Relational programming in Smalltalk

### 2019-03-14 booklet/article.tex
Describe the walking process and update the Chain hierarchy.

\subsection{Goals and substitutions}

@@

\subsection{Chains}

Because the sequence of substitutions may be infinite, we
represent it not as a \Verb|LinkedList| object but as a \Verb|Chain| object;
as usual, we find \Verb|BlockClosure| objects good candidates to represent (a possibly
infinite) computation. Some examples are in order,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest>>#fibs
    ^ self fib: 0 fib: 1

ChainTest>>#fib: m fib: n 
    ^ [self fib: n fib: m + n] links: m

ChainTest>>#gcd: m gcd: n 
    ^ [n isZero 
            ifTrue: [Chain with: m]
            ifFalse: [self gcd: n gcd: m \\ n]] 
        links: m

@@

ChainTest>>#ints: i 
    ^ [self ints: i + 1] links: i

ChainTest>>#upto2
    ^ [[[Chain bottom] links: 2] links: 1] links: 0
\end{minted}
where blocks define the message \Verb|#links:| as 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
BlockClosure>>#links: anObj
    ^ Chain item: anObj linker: self
\end{minted}
exposing the abstract class \Verb|Chain| and its class-side messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain class>>#bottom
    ^ Bottom new

@@

Chain class>>#with: anObj 
    ^ [Chain bottom] links: anObj

Chain class>>#repeat: anObj 
    ^ [self repeat: anObj] links: anObj
\end{minted}
together with its two subclasses \Verb|Knot| and \Verb|Bottom|; a comprehensive
test case follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest>>#testNumbers
    self 
        assert: Chain bottom contents 
        equals: LinkedList new.
    self
        assert: (self nats next: 10) contents
        equals: (0 to: 9).
    self
        assert: (self upto2 next: 10) contents
        equals: (0 to: 2).
    self
        assert: (self upto2 next: 1) contents
        equals: (0 to: 0).
    self 
        assert: (self fibs next: 10) contents 
        equals: {0. 1. 1. 2. 3. 5. 8. 13. 21. 34}.
    self 
        assert: ((self collatz: 10) next: 10) contents 
        equals: {10. 5. 16. 8. 4. 2. 1. 4. 2. 1}.
    self 
        assert: ((self gcd: 18 gcd: 32) next: 10) contents 
        equals: {18. 32. 18. 14. 4. 2. 2}.
    self 
        assert: ((Chain repeat: 4) next: 10) contents 
        equals: {4. 4. 4. 4. 4. 4. 4. 4. 4. 4}
\end{minted}
On one hand, when a \Verb|Chain| object (i)~receives the selector
\Verb|#next:| that carries a \Verb|SmallInteger| object $i$ it keeps at most
$i$ objects and then truncate itself, precisely  
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom>>#next: anInteger 
    ^ self

Knot>>#next: n 
    ^ n isZero
        ifTrue: [Chain bottom]
        ifFalse: [[self next value next: n - 1] links: item]
\end{minted}
On the other hand, when a \Verb|Chain| object (ii)~receives the selector
\Verb|#contents| it builds a \Verb|LinkedList| object \textit{eagerly},
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom>>#contents
    ^ LinkedList new

Knot>>#contents
    ^ self next contents 
        addFirst: item;
        yourself
\end{minted}
where \Verb|Knot| objects provide the following simple but fundamental message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Knot>>#next
    ^ linker value
\end{minted}
Observe that the \Verb|Chain| hierarchy support part of the protocol provided
by\Verb|ReadStream| objects because of the messages \Verb|#next|, \Verb|#next:|
and \Verb|#contents|.

>>>>

\subsection{\texttt{Goal}s and \texttt{State}s}

@@

\subsection{\texttt{Chain}s and \texttt{Cons}es}

Because the sequence of substitutions may be infinite, we
represent it not as a \Verb|LinkedList| object but as a \Verb|Chain| object;
as usual, we find \Verb|BlockClosure| objects good candidates to represent (a possibly
infinite) computation. Some examples are in order,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest>>#collatz: o
    ^ o
        cons: [o even
                ifTrue: [self collatz: o / 2]
                ifFalse: [self collatz: 3 * o + 1]]

ChainTest>>#fibs
    ^ self fib: 0 fib: 1

ChainTest>>#fib: m fib: n
    ^ m cons: [self fib: n fib: m + n]

ChainTest>>#gcd: m gcd: n
    ^ m cons: [n isZero
                ifTrue: [Chain with: m]
                ifFalse: [self gcd: n gcd: m \\ n]]

@@

ChainTest>>#ints: i
    ^ i cons: [self ints: i + 1]

ChainTest>>#zeroOneTwo
    ^ 0 cons: (1 cons: (2 cons: Chain bottom))
\end{minted}
where any \Verb|Object| object respond to the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object>>#cons: anObj
    ^ anObj consedObject: self
\end{minted}
which dispatches over
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object>>#consedObject: car
    ^ Cons car: car asCons cdr: self asCons

BlockClosure>>#consedObject: car
    ^ Chain item: car linker: self

Chain>>#consedObject: car
    ^ car cons: [self]
\end{minted}
where (i)~\Verb|Cons| objects mimic \textit{Lisp's cons cells}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cons class>>#car: anObj cdr: anotherObj 
    ^ self new 
        car: anObj;
        cdr: anotherObj;
        yourself

Object>>#asCons
    ^ Cons fromObject: self

Cons class>>#fromObject: anObj 
    ^ anObj
\end{minted}
and (ii)~the abstract class \Verb|Chain| provides the following class-side messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain class>>#bottom
    ^ Bottom new

@@

Chain class>>#with: anObj
    ^ anObj cons: Chain bottom

Chain class>>#repeat: anObj
    ^ anObj cons: [self repeat: anObj]
\end{minted}
together with its two subclasses \Verb|Knot| and \Verb|Bottom|. A comprehensive
test case follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest>>#testNumbers
    self
        assert: Chain bottom contents
        equals: LinkedList new.
    self
        assert: (self nats next: 10) contents
        equals: (0 to: 9).
    self
        assert: (self zeroOneTwo next: 10) contents
        equals: (0 to: 2).
    self
        assert: (self zeroOneTwo next: 1) contents
        equals: (0 to: 0).
    self
        assert: (self fibs next: 10) contents
        equals: {0. 1. 1. 2. 3. 5. 8. 13. 21. 34}.
    self
        assert: ((self collatz: 10) next: 10) contents
        equals: {10. 5. 16. 8. 4. 2. 1. 4. 2. 1}.
    self
        assert: ((self gcd: 18 gcd: 32) next: 10) contents
        equals: {18. 32. 18. 14. 4. 2. 2}.
    self
        assert: ((Chain repeat: 4) next: 10) contents
        equals: {4. 4. 4. 4. 4. 4. 4. 4. 4. 4}
\end{minted}

On one hand, when a \Verb|Chain| object (i)~receives the selector
\Verb|#next:| that carries a \Verb|SmallInteger| object $i$ it keeps at most
$i$ objects and then truncate itself, precisely
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom>>#next: anInteger
    ^ self

Knot>>#next: n
    ^ n isZero
        ifTrue: [Chain bottom]
        ifFalse: [[self next value next: n - 1] links: item]
\end{minted}

On the other hand, when a \Verb|Chain| object (ii)~receives the selector
\Verb|#contents| it builds a \Verb|LinkedList| object \textit{eagerly},
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom>>#contents
    ^ LinkedList new

Knot>>#contents
    ^ self next contents
        addFirst: item;
        yourself
\end{minted}
where \Verb|Knot| objects provide the following simple but fundamental message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Knot>>#next
    ^ linker value
\end{minted}
Observe that the \Verb|Chain| hierarchy support the protocol provided
by \Verb|ReadStream| objects because of the messages \Verb|#next|, \Verb|#next:|
and \Verb|#contents|.

\subsection{\texttt{Runner}s, \texttt{Reifier}s and \texttt{Var}s}

Recall that a goal $g$ responds with a \Verb|Chain| that contains either zero,
one, or more substitutions when it receives the message \Verb|#onState:|; in order
to manipulate those substitutions, $g$ provides the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#solutions
    ^ Runner onGoal: self

Runner class>>#onGoal: aGoal 
    ^ self new 
        goal: aGoal;
        yourself
\end{minted}
which builds a \Verb|Runner| object able to process the sequence of substitution, both (i)~all of them with the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#all
    | chain |
    chain := goal onState: self emptyState.
    ^ self reify: chain contents
\end{minted}
and (ii)~at most $n$ of them with the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#atMost: anInteger 
    | chain |
    chain := goal onState: self emptyState.
    ^ self reify: (chain next: anInteger) contents
\end{minted}
using the helper message 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#emptyState
    ^ Dictionary new asState
\end{minted}

We are now ready to describe the reification process which starts in
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#reify: states 
    | vars |
    vars := goal vars ifEmpty: [{Var tautology}].
    ^ states
        collect: [:s | "`s` stands for `state`"
                    | rs | "`rs` stands for `reified state`"
                    rs := s reify: vars.
                    rs size = 1 ifTrue: [rs first] ifFalse: [rs]]
\end{minted}
where the accessor \Verb|#vars| is defined for any \Verb|Goal| objects and overridden by \Verb|Fresh| goals 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#vars
    ^ {}

Fresh>>#vars
    ^ vars
\end{minted}
respectively; in parallel, every \Verb|State| object responds to the selector \Verb|#reify:| with
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#reify: vars 
    | walker reifier |
    walker := Walker on: self.
    reifier := Reifier on: Dictionary new asState.
    ^ vars
        collect: [:v | 
                    | w s |
                    w := walker walkStar: v.
                    s := reifier reify: w.
                    (Walker on: s) walkStar: w]
\end{minted}
where a \Verb|Walker| object is built by 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Walker class>>#on: aState
    ^ self new 
        state: aState;
        yourself
\end{minted}
and it responds to the selector \Verb|#walkStar| with
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Walker>>#walkStar: aVar 
    ^ (state walk: aVar)
        walkStarUsingWalker: self
\end{minted}
which relies on 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#walk: anObj 
    | k |
    k := anObj.
    [k := substitution
            at: k
            ifAbsent: [^ k]] repeat
\end{minted}
and different objects responds to \Verb|#walkStarUsingWalker:| as follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object>>#walkStarUsingWalker: aWalker
    ^ aWalker walkStarObject: self

Array>>#walkStarUsingWalker: aWalker
    ^ aWalker walkStarArray: self

LinkedList>>#walkStarUsingWalker: aWalker
    ^ aWalker walkStarLinkedList: self

Cons>>#walkStarUsingWalker: aWalker
    ^ aWalker walkStarCons: self 
\end{minted}
dispatching over \Verb|Walker| objects
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Walker>>#walkStarObject: anObj 
    ^ state walk: anObj

Walker>>#walkStarArray: aCollection 
    ^ aCollection
        collect: [:each | self walkStar: each]

Walker>>#walkStarLinkedList: aCollection 
    ^ aCollection
        collect: [:each | self walkStar: each]

Walker>>#walkStarCons: aCons 
    | car cdr |
    car := self walkStar: aCons car.
    cdr := self walkStar: aCons cdr.
    ^ car cons: cdr 
\end{minted}
respectively.
